exponential Bradley-Terry

### 2025-10-07 inst/exponential_BT.Rmd
simple cases

>>>>

# Some simple cases

We consider the smallest non-trivial cases $n=3$ and $n=4$.  For $n=3$
we have Plackett-Luce strengths $\alpha x,\alpha x^2,\alpha x^3$ [the
  extra factor of $x$ allows us to identify the $r$ in ``$\alpha x^r$''
  with the rank of the competitor, the best being rank 1].  Suppose
the focal competitor came first, the likelihood function would be

\begin{equation}
  \mathcal{L}(x,r)=\frac{x^{r-1}}{1+x+x^2}\qquad r=1,2,3
\end{equation}

```{r plot123first}
f <- function(x,r){ x^r/(x + x^2 + x^3) }

x <- seq(from=0.01 , to=1, by=0.01)
plot(c(-3,0),c(-3,1),type='n', xlab="log x", ylab="support",
     main = "support for x,r", sub="Three competitors, focal competitor came FIRST"
     )
points(log(x),log(f(x,1)),type="l",col="black")
points(log(x),log(f(x,2)),type="l", col="red")
points(log(x),log(f(x,3)),type="l", col="blue")

abline(h=c(0,-2))

legend("topright",lty=1, col=c("black","red","blue"),
       legend=c("r=1","r=2","r=3"))

```

If the focal competitor came second we would have two Plackett-Luce
probabilities to add, corresponding to different order statistics.
$\alpha x^i$ or $\alpha x^j$ coming first where $i,j\neq r$.  For
example, if $r=1$ we would have two finishing orders that are
consistent with the observation: $2\succ 1\succ 3$ and $3\succ 1\succ
2$.  These are disjoint so the likelihood will be proportional to

$$
\frac{x^2}{x+x^2+x^3}\cdot\frac{x}{x+x^3} + 
\frac{x^3}{x+x^2+x^3}\cdot\frac{x}{x+x^2}
$$

After simplification:

\begin{equation}
  \mathcal{L}(x,r)=
  \begin{cases}
    \frac{x  }{x+x^2+x^3}\left(\frac{x^2}{x+x^3}   + \frac{x^3}{x+x^2}\right)\qquad r=1\\
    \frac{x^2}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^3}{x+x^2}\right)\qquad r=2\\
    \frac{x^3}{x+x^2+x^3}\left(\frac{x  }{x^2+x^3} + \frac{x^2}{x+x^3}\right)\qquad r=3
\end{cases}
\end{equation}

```{r plot123second}
like <- function(x,r){
  S <- x + x^2 + x^3
  switch(r,
         '1' = x^2/S*x^1/(x^1+x^3) + x^3/S*x^1/(x^1+x^2), # 213+312
         '2' = x^1/S*x^2/(x^2+x^3) + x^3/S*x^2/(x^1+x^2), # 123+321
         '3' = x^1/S*x^3/(x^2+x^3) + x^2/S*x^3/(x^1+x^3)  # 132+231
  )
}

x <- seq(from=0.01 , to=1, by=0.01)
plot(c(-3,0),c(-3,1),type='n', xlab="log x", ylab="support",
     main = "support for x,r", sub="Three competitors, focal competitor came SECOND"
     )
points(log(x),log(like(x,1)),type="l",col="black")
points(log(x),log(like(x,2)),type="l", col="red")
points(log(x),log(like(x,3)),type="l", col="blue")

abline(h=c(0,-2))

legend("topright",lty=1, col=c("black","red","blue"),
       legend=c("r=1","r=2","r=3"))

```

Now the focal competitor comes last:

```{r plot123third}
like <- function(x,r){
  S <- x + x^2 + x^3
  switch(r,
         '1' = x^3/S*x^2/(x^1+x^2) + x^2/S*x^3/(x^1+x^3), # 213+312
         '2' = x^1/S*x^3/(x^2+x^3) + x^3/S*x^1/(x^1+x^2), # 123+321
         '3' = x^1/S*x^2/(x^2+x^3) + x^2/S*x^1/(x^1+x^3)  # 132+231
  )
}

x <- seq(from=0.01 , to=1, by=0.01)
plot(c(-3,0),c(-3,1),type='n', xlab="log x", ylab="support",
     main = "support for x,r", sub="Three competitors, focal competitor came THIRD"
     )
points(log(x),log(like(x,1)),type="l",col="black")
points(log(x),log(like(x,2)),type="l", col="red")
points(log(x),log(like(x,3)),type="l", col="blue")

abline(h=c(0,-2))

legend("topright",lty=1, col=c("black","red","blue"),
       legend=c("r=1","r=2","r=3"))

```

The $n=4$ case is harder as we potentially have six consistent
orderings for each observation.  As an example, if the focal
competitor comes third we have $1
